Survival in chessland

### 2018-12-15 trunk/chess/papers/survival.tex
Improvements to survival table, paper text

@@

@@

If you are forced to play chess to the death, you are in trouble,
because most people are not good at chess (for example, the author)
and yet want to live.\footnote{It is easy for two players to
  collaborate to produce a draw, especially by simply agreeing to a
  draw at the outset of the game (if allowed). Some tournament formats
  forbid the players from agreeing to a draw verbally before a certain
  point (e.g. 30 moves). There are always other routes to a draw, for
  example by stalemate or repeating the same position three times.
  Collaboratively producing such situations is easy, but this strategy
  is not likely a stable equilibrium: Players can often gain a
  substantial advantage by going ``off script'' and instead trying to
  win the game. Additionally, sometimes the terms of
  chess-to-the-death do not allow the players to communicate at all
  beforehand, nor during the game. If this is the case, then it may be
  difficult to agree on the approach to drawing, let alone establish
  that this is both player's desire. Since the rules of
  chess-to-the-death can't forbid us from colluding at this moment, I
  hereby declare that the following is the correct approach: 
%
  1. Nf3. This is a reasonable opening move for white (``Zukertort opening'')
  which can transpose into several common systems (e.g. King's
  Indian). Since the knight can move back to g1 on the next move,
  knight moves are the fastest route to a draw by repetition. This
  has a good chance of signaling to a wise player that a draw is
  desired. The player should make this move after pondering carefully
  for some time, and then looking meaningfully into the other player's
  eyes. 
  1. ... Nf6. This is both a strong response for black in a real game,
  and simultaneously a signal that a draw is desired. The other
  advantage is that very weak players may simply copy what white does.
  In doing so, they will also play this move. 
  2. Ng1?!. This is a terrible move for white, but clearly signals
  the intention to draw.
  2. ... Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
  for a second time.
  3. Nf3 Nf6. At this point the game should clearly continue repeating
  the sequence, although since we are in the start position, white has
  any number of strong opening moves available. Signaling the draw line
  and then 3. d4!? may be pyschologically devastating.
  4. Ng1 Ng8 1/2-1/2. The starting position is reached for the third
  time, which by rule is a draw.

@@

  If 1. ... Nc6 or another Knight's move, white can also consider
  continuing in the obvious way. However after 1. ... d5, black
  has refused or not noticed the draw. Fortunately, white is still
  in a good position to play the game normally. White can try
  to be more obvious with 2. Ng1, but if black is choosing to just
  play normally, white takes a distinct handicap by doing so.

@@

  Of course, this all relies on the assumption that if
  chess-to-the-death ends in a draw, the players are spared or allowed
  to repeat indefinitely. If both players are actually executed, then
  this line is truly a Fool's Draw! }
%
But what if you are forced to {\it be one of the chess pieces} to the
death? That is, your little soul inhabits one of the 32 pieces or
pawns and your soul is vanquished if that piece is eliminated. Now it
doesn't matter whether you're good or bad at chess, because you don't
get to pick what happens in the game. What matters is that your piece
lives to the end of the game, when all surviving pieces are set free.
Which piece should you want to be?

@@

Like all good scientific research, I clearly laid out my hypothesis
and wrote down the motivation before performing the study. This helps
prevent presentation bias where the results appear more satisfactory
because they are framed as a natural conclusion from the idea that
motivated the research in the first place (when in fact, of course,
if you write the motivation after witnessing the results, ..). 
It is also much more exciting. I literally don't know the answer
as I'm writing this, nor whether it is interesting in any way!

@@

\subsection{Ben}

@@

5. The C,D,E, F pawns
 -- Moved forward to release various more important pieces => more likely to die.

6. Both Bishops
 -- https://www.youtube.com/watch?v=gDnE-5lD7w8

@@

>>>>

% \usepackage{chessfs}
\usepackage{adjustbox}

% Define black versions of pieces for inline use. Gross, but it works.
\newcommand{\Pawn}[1][1.3ex]{%
\adjustbox{Trim=4.3pt 2.6pt 4.3pt 0pt,width=#1,margin=0.2ex 0ex 0.2ex 0ex}{\BlackPawnOnWhite}%
}%
\newcommand{\Rook}[1][1.58ex]{%
\adjustbox{Trim=3.2pt 2.2pt 3.2pt 0pt,width=#1,raise=0ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackRookOnWhite}%
}%
\newcommand{\Knight}[1][1.85ex]{%
\adjustbox{Trim=2.3pt 2.35pt 2.5pt 0pt,width=#1,raise=-0.03ex,margin=0.14ex 0ex 0.14ex 0ex}{\BlackKnightOnWhite}%
}%
\newcommand{\Bishop}[1][1.79ex]{%
\adjustbox{Trim=2.3pt 2pt 2.3pt 0pt,width=#1,raise=-0.12ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackBishopOnWhite}%
}%
\newcommand{\Queen}[1][2.05ex]{%
\adjustbox{Trim=1.2pt 2.2pt 1.2pt 0pt,width=#1,raise=-0.08ex,margin=0.1ex 0ex 0.1ex 0ex}{\BlackQueenOnWhite}%
}%
\newcommand{\King}[1][1.95ex]{%
\adjustbox{Trim=2pt 2pt 2pt 0pt,width=#1,raise=-0.06ex,margin=0.13ex 0ex 0.13ex 0ex}{\BlackKingOnWhite}%
}%

% lets me explicitly set a. or 1. etc. as enum label
\usepackage{enumitem}

@@

\setlength{\footnotesep}{2em}

@@

If you are forced to play chess to the death, you are in trouble,
because most people are not good at chess (for example, the author)
and yet want to live.\footnote{It is easy for two players to
  collaborate to produce a draw, especially by simply agreeing to a
  draw at the outset of the game (if allowed). Some tournament formats
  forbid the players from agreeing to a draw verbally before a certain
  point (e.g. 30 moves). There are always other routes to a draw, for
  example by stalemate or repeating the same position three times.
  Collaboratively producing such situations is easy, but this strategy
  is not likely a stable equilibrium: Players can often gain a
  substantial advantage by going ``off script'' and instead trying to
  win the game. Additionally, sometimes the terms of
  chess-to-the-death do not allow the players to communicate at all
  beforehand, nor during the game. If this is the case, then it may be
  difficult to agree on the approach to drawing, let alone establish
  that this is both player's desire. Since the rules of
  chess-to-the-death can't forbid us from colluding at this moment, I
  hereby declare that the following is the correct approach: 
\begin{enumerate}[label=\arabic*.]
  \item[1.] Nf3. This is a reasonable opening move for white (begins the R\'eti)
    which can transpose into several common systems (e.g. King's
    Indian). Since the knight can move back to g1 on the next move,
    knight moves are the fastest route to a draw by repetition. This
    has a good chance of signaling to a wise player that a draw is
    desired. The player should make this move after pondering carefully
    for some time, and then looking meaningfully into the other player's
    eyes. 
  \item[1.] ... Nf6. This is both a strong response for black in a real game,
    and simultaneously a signal that a draw is desired. The other
    advantage is that very weak players may simply copy what white does.
    In doing so, they will also play this move. 
  \item[2.] Ng1?!. White moves the knight back to its starting square.
    This is a terrible move for white, but clearly signals the intention to
    draw.
  \item[2.] ... Ng8!. ``Fool's Draw Accepted.'' The starting position is reached
    for a second time.
  \item[3.] Nf3 Nf6. At this point the game should clearly continue repeating
    the sequence, although since we are in the start position, white has
    any number of strong opening moves available. Signaling the draw line
    and then 3. d4!? may be pyschologically devastating.
  \item[4.] Ng1 Ng8 1/2-1/2. The starting position is reached for the third
    time, which by rule\footnotemark\ is a draw.
\end{enumerate}

@@

  If 1. ... Nc6 or another Knight's move, white can also consider
  continuing in the obvious way. However after 1. ... d5, black has
  refused or not noticed the draw. Fortunately, white is still in a
  good position to play the game normally (this is the main line of
  the R\'eti opening, followed by 2. c4). White can try to be more
  obvious with 2. Ng1, but if black is choosing to just play normally,
  white takes a distinct handicap by doing so.

@@

  Of course, this all relies on the assumption that if
  chess-to-the-death ends in a draw, the players are spared or allowed
  to repeat indefinitely. If both players are actually executed, then
  this line is truly a Fool's Draw!
}
%
\footnotetext{
  But is it? The lichess implementation (although known
  to be buggy\cite{kingme}) does not permit a threefold repetition
  claim in this situation, which got me thinking that maybe there is
  some subtlety here. Is the starting position special somehow, not
  counting as having occurred? The relevant statute, from the FIDE
  Laws of Chess\cite{fiderules}:
  \begin{quote}
    {\bf 9.2}.\quad The game is drawn, upon a correct claim by a
    player having the move, when the same position for at least the
    third time (not necessarily by a repetition of moves):
    \begin{enumerate}[label=\alph*.]
    \item is about to appear, if he first writes his move, which
      cannot be changed, on his scoresheet and declares to the arbiter
      his intention to make this move, or
    \item has just appeared, and the player claiming the draw has the
      move.
  \end{enumerate}

  Positions are considered the same if and only if the same player has
  the move, pieces of the same kind and colour occupy the same squares
  and the possible moves of all the pieces of both players are the
  same. Thus positions are not the same if:
  \begin{enumerate}[label=\arabic*.]
  \item at the start of the sequence a pawn could have been captured
    en passant.
  \item a king or rook had castling rights, but forfeited these after
    moving. The castling rights are lost only after the king or rook
    is moved.
  \end{enumerate}
  \end{quote}

  So the question is, has the starting position ``appeared'' before
  white's first move? The rules are not totally clear on this point.
  Note that ``positions are considered the same'' only when the
  same player ``has the move.'' FIDE defines ``have the move'' as
  \begin{quote}
    1.3.\quad A player is said to 'have the move' when his opponent's
    move has been 'made'.
  \end{quote}
  A strong case can therefore be made that white does not 'have the
  move' in the formal sense at the beginning of the game, since black
  has not made a move!

  Nonetheless, it does seem clear that white can claim a draw by
  9.2.a, by committing the move 5. Nf3 and declaring to the arbiter
  that the position is now {\it about to appear} for the third time.
  This seems unambiguously legal. }
%
But what if you are forced to {\it be one of the chess pieces} to the
death? That is, your little soul inhabits one of the 32 pieces or
pawns and your soul is vanquished if that piece is eliminated. Now it
doesn't matter whether you're good or bad at chess, because you don't
get to pick what happens in the game. What matters is that your piece
lives to the end of the game, when all surviving pieces are set free.
Which piece should you want to be?

@@

Like all good scientific research, I clearly laid out my hypothesis
and wrote down the motivation before performing the study. This helps
prevent presentation bias where the results appear more satisfactory
because they are framed as a natural conclusion from the idea that
motivated the research in the first place (when in fact, of course, if
you write the motivation after witnessing the results, backflow is
inevitable). It is also much more exciting. I literally don't know the
answer as I'm writing this, nor whether it is interesting in any way!

@@

\subsection{Ben}\footnote{Ben does not prefer to use the shift key.}

@@

5. The C,D,E,F pawns
 -- Moved forward to release various more important pieces $\Rightarrow$ more likely to die.

6. Both Bishops
 -- https://youtu.be/gDnE-5lD7w8

@@

\section{Methodology}

To compute the chances for survival, I legally acquired 506,000,416
chess games from lichess.org. This is all of the standard variant,
rated games from Jan 2013 to November 2018, in any time format. Only
games that are completed and valid are included (about 200,000 games
did not meet this criteria). The total data size is 875 gigabytes, so
processing these took some care for efficiency and parallelism.

Other than that, I simply parsed and simulated each game. For each
of the 32 pieces in the starting position, I tracked its current
location, and whether it is alive; multiple dead pieces could
occupy the same square. At the end of the game, one of the kings is
killed if his side has lost.

For a piece, there is a single factual survival rate in these games,
given just by $\frac{\mathrm{num~survived}}{\mathrm{num~games}}$. What
we're really interested in, however, is estimating the underlying true
survival probability for each piece. In order to do this with
reasonable efficency, we divide the games into 32 separate buckets,
and count statistics separately for each. From these samples we can
then estimate variance, for example. The games are bucketized by a
deterministic hash of the White player's username. This way, if there
exist some players who are highly unusual (perhaps automated
accounts), their games are grouped together and pessimistically
represented in the variance estimate. This also helps account for
different opening preferences; the chosen opening certainly affects
the survival chances.

The basic survival chances appear in Figure~\ref{fig:piece-survival}.
Indeed, many pieces are more likely to survive than the kings. Even as
black, the extremal pawns (\pawn a and \pawn h) have over a 70\%
survival rate. Across the board, the survival chances for a white
piece and its black twin are similar, usually with a small edge to
white. Notable exceptions are the \knight g (the overall most doomed
piece), and both white bishops, which die more than their black
counterparts. The \pawn e is vastly more dead than \Pawn e. Note
somewhat satisfyingly that the \pawn c has the highest variance; this
was the most controversial among the drinking buddies
(Section~\ref{sec:XXXbuddies}). Note that \pawn c is the sacrificed pawn
  in the popular Queen's Gambit (1. d4 d5 2. c4); variance may come
  from individual players having a strong preference to accept or
  decline.\footnote{XXX actually this can't be true because we group
    by the white player, but the black player chooses what to do.}

% XXX

\begin{figure}[htb]
  \begin{center}
    % To generate this pdf, I ran maketable to make
    % piece-survival.svg, then printed to PDF using Chrome, then
    % cropped it in Illustrator. Illustrator can open the SVG but
    % doesn't have the unicode chess pieces in any font I could find.
    \includegraphics[width=0.95 \linewidth]{piece-survival-export}
  \end{center}\vspace{-0.1in}
  \caption{Survival probabilities for each of the 32 pieces in standard
    chess, in 500 million games. The number is the mean survival rate
    across all samples. The vertical position of the dot is this mean
    rate, with a line drawing the span between the smallest and largest
    sample bucket (this is usually a very tiny range).
    Horizontal position is purely presentational, to avoid overlap.}
  \label{fig:piece-survival}
\end{figure}
